\textbf{Beasts of Prey or Rational Animals?\\ Private Governance in Brazil's \textit{Jogo do Bicho}}

### 2018-11-18 paper-bicho/response-reviewers.tex

\noindent 1) \textit{In one case the vagueness of the theory is clear ``civil servants can be co-opted by criminal groups and produce sub-optimal social outcomes.'' Sub-optimal is vague. [\dots] The problem I have with the theoretical approaches used in this text is that they are neutralistic, evading discussions of social conflict in favor of an amalgam of theories based on rational choice and functional aspects of government.}

\vspace{.25cm}

I completely agree with the comment. I have made substantial changes to the manuscript to better illustrate the social tensions inherent in the \textit{jogo do bicho}. For instance, in section 2.2 I highlight that ``[\dots] \textit{the banqueiros do not use violence only against other criminals. They often employ violent methods against competitors and their own staff, too}." Then I give examples of how the \textit{jogo do bicho} elites frequently exploit their subordinates. Similarly, in section 4.2 I describe other ways by which the \textit{banqueiros} enforce their rule over employees and members of civil society. In this sense, social conflicts are more prominent in my description of the lottery.

I also adopt a more nuanced approach when analysing the relationship between the \textit{jogo} and the public. In the last 5 paragraphs of section 2.2, I point out that not only the animal game has negatively affected the formal labour market, but it has also reduced the demand for social rights in Brazil. The \textit{jogo} provides jobs for those would could not easily enter the labour force, and as result ``[\dots] \textit{poor workers are able to consume without resorting to government assistance, so state officials can spend a larger amount of public funds spent on other, probably wealthier, sectors of the population}.'' I then mention how the \textit{jogo} has collaborated to increase inequality levels in the country, mainly by transferring resources to rich animal game members and restricting access to newcomers. The game elites frequently employ violent methods to secure their rents against demands for redistribution from employees. Discussions about how the animal game distorts the allocation of public funds and corrupts political representation are also mentioned in section 2.2.

I believe that my analysis of the text is not as neutralistic as before, and hopefully these changes have illustrated how pervasive are social tensions in the \textit{jogo do bicho}.

@@

\vspace{.5cm}

\noindent 2) \textit{I think this could have been a more LAP style paper if it had expanded on the role of gambling rackets in the economy, the specific types of political candidates supported by the jogo, the overall impact of this much cash circulating and how it is harnessed by the formal economy, and the manipulation of poor district votes. I find much of the discussion on how the game is perceived and the theories of choice to be less interesting. What welfare provisions are provided by the jogo? Why is the only discussion of violence saved for the conclusion? The most interesting questions (as far as the VCC issue is concerned) are found in the conclusion.}

\vspace{.25cm}

The reviewer is correct to point out that I should have described those topics should have in further detail. I now provide a detailed discussion about the role of gambling rackets in Brazil's formal economy. Some of the paragraphs I have added to the revised manuscript address specifically the issue raised by the reviewer. In the case of the political candidates supported by the \textit{jogo}, I note in paragraph 3 of section 4.2 that `[\dots] \textit{Politicians from all spheres of government are involved with jogo do bicho bankers. Recent investigations have shown that from local representatives to senators, politicians of every level receive illegal money to fund their campaigns.}.' The claim is backed by a description of Carlinhos Cachoeira's wide political patronage network. Cachoeira is a well-known \textit{jogo do bicho} boss in his home state of Goiás. Additionally, as I mentioned in the previous response, I discuss the overall impact of the \textit{jogo} in the formal economy in the last paragraphs of section 2.2.

@@

\noindent 1) \textit{It is written in (British) English, and there are minor grammatical problems that can be easily corrected. It should not be too difficult to edit. Also, though the actual article is 27 pages, the extensive bibliography (mostly made up of secondary sources) takes up another 19 pages! I wonder how much of this is necessary.}

\vspace{.25cm}

I would like to express my gratitude to Paulo Simões for his excellent comments, which encouraged me to improve the article along the lines suggested.

The article has been proofread and the text now uses American spelling. With regards to the reference list, I fully agree that it was too long. The bibliography is now about 50\% shorter, and in my view the text remains as informative as before. Hopefully the manuscript is now easier to read.

@@

\vspace{.25cm}

I have included several newspaper stories to the article with the aim of making it more appealing to a wider audience. For instance, in section 2.2 I mention a story about two bankers and their rise to power in Rio de Janeiro:

\begin{quote}
	\citet{jupiara2015poroes} argue that Ailton Guimarães Jorge, a former Army officer, tortured and murdered rival lottery bosses in the late 1970s. One of his former allies, Army Colonel Paulo Malhães, told to the Rio de Janeiro State Truth Commission that Guimarães `went on a rampage' to consolidate his power \citep{belem2015guimaraes}. Castor de Andrade, Rio's most influential animal game banker, also employed similar methods to run his business. Andrade kept an armed bodyguard of 23 men and allegedly murdered a number of competitors. In a famous case, Andrade shot Euclides Ponar, an old jogo do bicho boss known as `Grey-Headed Chinese' (\textit{China Cabeça Branca}) after Ponar denounced a fraud in lottery draws in 1976. Andrade was likely involved in other assassination plots in the 1990s \citep{globo2017castor}.
\end{quote}

There are similar excerpts in other parts of the manuscript. For instance, I add some information about the commercialisation of slot machines by \textit{jogo do bicho} members:

@@

\subsection*{Reviewer 3: Matthew Lorenzen}%
\label{sub:reviewer_3_matthew_lorenzen}

\noindent 1) \textit{The author's methodology is, at first, not clear. It eventually becomes apparent that the article is based on an analysis of bibliographical sources, and not on any primary data or secondary statistical data. The author should make this clear in the introduction.}

\vspace{.25cm}

Thank you very much for your comments. The introduction now states that the article uses secondary sources exclusively. Indeed this was not obvious to readers and the reviewer was correct to ask for further clarifications. The introduction now reads as follows:

@@

\vspace{.5cm}

\noindent 3) \textit{In addition, the long examination of the historical background, including the directed acyclic graph, does not seem to be crucial for the rest of the article. Moreover, by attempting to address too many issues, the author does not focus on a theoretical perspective (or contrasts theoretical perspectives), but instead briefly mentions different perspectives for each separate issue (repugnant markets, game theory, state capture...). I suggest the author to focus more keenly on one or two issues and one or two different theoretical perspectives, rather than attempting to address all topics in an overarching way.}

\vspace{.25cm}

This is an important point. The article deals with a number of topics that at first seemed only loosely associated. The section on repugnance, for instance, added little to the main ideas of the text and is not required to understand the remaining of the article. That section has been removed and the subsections on costly signalling and reputation strategies are longer. The article is, in my view, more focused and concise as a result.

@@

\subsection*{Reviewer 4: Andrew Smolski}%
\label{sub:reviewer_4_andrew_smolski}

\noindent 1) \textit{Agents are differentiated by their role (bookmaker, banker, etc.), yet there is little discussion of how they form a hierarchy of power to affect. It appears as if each agent has a similar power to affect, which muddles the understanding of the ability of bicheiros to implement their strategies. For instance, in the discussion of state capture through the samba schools, a prior understanding of the hierarchy of roles would help the reader to understand the ability to impose oligopolistic conditions, as well as strengthen the argument concerning the role of bicheiros in market formation. See: Neil Fligstein's work on incumbent-challengers in capitalist markets, a political-cultural theory of market formation and alteration}.

\vspace{.25cm}

The referee rightly points out that I did not emphasise the power structures of the animal game strongly enough. This criticism also appeared in other referee reports and I am happy to address it. As previously mentioned, section 2.2 has been largely rewritten and it now includes new paragraphs on the social conflicts within the \textit{jogo}. I have also changed the section title to `A Hierarchical Organisational Structure' to make this point clearer. The indication of Neil Fligstein's work is also very appropriate and I sincerely thank the referee for it. I have added one paragraph about Fligstein's concept of `markets as politics' to conclude section 2.2.

@@

\vspace{.5cm}

\noindent 2) \textit{Further, and following from this concern over hierarchy, at the beginning of section 2.1, the author writes, \emph{``Spontaneous orders are emergent macro-level phenomena that result from voluntary actions of purposive, self-interested individuals utilizing their contextual knowledge}'' (emphasis mine). What does the author mean by voluntary? For instance, urban poverty as a structural factor places severe limitations on what actions can be taken, and this goes beyond just a matter of knowledge. So, while neo-institutionalists like North would reduce institutions to schemas for action, others in the institutional political economy and neo-Marxist traditions, argue that institutions are also very much a matter of power (see Portes 2010 for instance). Considering LAP is a critical journal, the author would do well to explain the rational choice sense of agency, and contrast it with more critical accounts.}

\vspace{.25cm}

In the revised version of the manuscript, I have removed the first part of section 2 -- the one quoted by the reviewer -- because I do agree that it is misleading. The quote downplays the violent aspects of the animal game's internal governance, and as such I do not believe it should be included. In contrast, I have added examples of how the \textit{jogo} bosses employ violence, or the threat thereof, against their own employees to enforce compliance to their decisions.

Nevertheless, in the article I argue that violence can be used rationally, and I believe it is the case with the \textit{jogo}. However, I adopt a model of rationality which also assumes humans are embedded in a myriad of formal and informal institutions that design and shape their choices. I concur with the reviewer that no individual is free of social constrains. It is indeed unrealistic to assume otherwise. Apart from institutional limits, individuals have limited access to information, what reduces their abilities to correctly evaluate alternative scenarios. But in the \textit{jogo}, individuals generally behave in ways that maximise their own interest, so I believe rationally is not fully absent from the dynamics of the game. This idea of `thin rationality' conforms well to a wide range of the literature on crime and violence, such as \citet{balcells2010rivalry}, \citet{gambetta1996sicilian,gambetta2009codes}, \citet{kalyvas2006logic}, \citet{leeson2010pirational}, or \citet{skarbek2011governance,skarbek2012prison}.

I did not include a lengthier discussion about the rational choice sense of agency because I thought it would be slightly too abstract for my purposes. The article is mainly focused on the bankers' strategies for long-term survival, and I believe readers will be more interested in more concrete aspects of the animal game. But I did add a footnote to the introduction describing the notion of rationality I employ in the text. The footnote reads as follows:

@@

	[\dots] For instance, the \textit{jogo do bicho} gives a boost to the Brazilian economy by providing jobs for unskilled workers who cannot easily join the labour market. By doing so, the \textit{jogo} prevents some of poorest members of the Brazilian society from demanding more inclusive government policies, although they are the ones who would benefit the most from public assistance. Because of the income generated by the \textit{jogo do bicho}, poor workers are able to consume without resorting to government assistance, so state officials can spend a larger amount of public funds spent on other, probably wealthier, sectors of the population.

	[\dots] Another indirect economic effect of the \textit{jogo} is the rise in inequality. While formal workers can -- at least in theory -- demand higher compensations during a market upswing, the same is not true for the animal game employees. The threat of violence reduces the space for collective bargaining with the lottery bosses, and higher profits at the top of the \textit{jogo do bicho} structure do not trickle down to those at the bottom. This is one of the reasons why the game bankers turned the game into an oligopoly: the use of violence guarantees not only that new entrants will not be allowed to join the market, but also that profits are concentrated in the hands of very few individuals.  
\end{quote}

\vspace{.5cm}

\noindent 4) \textit{The point made by Arrow used in section 3.3 that criminals ``do not put the required levels of effort if they know they are not being monitored'', lacks empirical backing in the literature. It is to take a doxa about criminals and treat it as true. I suggest looking at contemporary work on pirates for instance, and other extra-legal organizations, instead of using moral judgements as theoretical arguments. Remember, who is and who is not a criminal is a social matter, not a universal. Police invading a favela is a criminal act, but we do not label them criminals. Or, we can state that criminals are always those that follow Arrow's definition, but then you would have to demonstrate that the agents in the jogo do bicho act this way. Otherwise, it substitutes empirical reality with a possibly unfounded theoretical conjecture. This also returns us to point 3, whereby clarifying extra-legal economy would help to understand extra-legal actors.} 
\vspace{.25cm}

I think this is an important criticism. Due to my poor phrasing, the sentence highlighted by the referee does suggest a moralistic undertone. Yet I would like to affirm that it was not the case. My intention was not to use a moral judgements to describe the character or motivation of individuals, but only to draw attention to potential agent-principal dilemmas in the game. This problem is widely discussed in the economics literature and it is not ascribed only to informal or illegal activities. Agency dilemmas are pervasive in firms big or small and in government agencies too \citep[e.g.,][]{grossman1983analysis,laffont2009theory,sappington1991incentives}. I sincerely apologise for the confusion my statement has caused.

I have removed the mention to Arrow's work to avoid further misunderstandings. I believe the revised manuscript exposes more clearly the social tensions of the animal game, and this has improved the quality of the text.

### 2018-11-18 paper-bicho/bicho.tex
proofreading done

\begin{document}
\maketitle
\begin{abstract}
\noindent
This article presents a rational choice account of Brazil's \textit{jogo do bicho} (`animal game'), possibly the largest illegal lottery game in the world. Over 120 years, the \textit{jogo do bicho} has grown from a local raffle to a multimillion-dollar business, and the game has played a major role in Brazil's cultural and political life. My analysis of the \textit{jogo do bicho} has three goals. First, I examine the mechanisms that fostered the lottery's notable growth outside of the boundaries of Brazilian law. Second, I investigate how the animal game financiers combine costly signals and selective incentives to induce cooperation from members of community. Lastly, I discuss the relationship between the lottery sponsors and Brazilian representatives, particularly how the \textit{bicheiros} exploit the fragmentation of Brazil's political system to advance their long-term interests.

@@

In 1892, Baron João Batista de Viana Drummond came up with a new idea to fund his cash-strapped zoo. Situated in a quiet neighbourhood in the north of Rio de Janeiro, the \textit{Jardim Zoológico}, or Zoological Garden, hosted a variety of exotic species and offered breath-taking views of the city. But it lacked visitors. An experienced businessman, Drummond realised the zoo would have to provide other kinds of entertainment to keep itself afloat. One of his plans seemed particularly promising: a lottery raffle.

The rules were straightforward. In the morning, the Baron would choose one animal from a list of 25 beasts and put its picture inside a wooden box at the zoo's entrance. Visitors who wanted to join the raffle received a ticket bearing the stamp of one of those 25 animals.\footnote{At first, the zoo staff distributed the tickets at random, so the game consisted in a simple raffle. However, the zoo soon allowed participants to choose the animals they preferred. This small change made the game considerably more appealing as it introduced an element of divination to the \textit{jogo} \citep[71--74]{da1999aguias}.} The lucky winner would take home a prize worth 20 times the ticket price, an amount higher than a carpenter's monthly wage \citep[542]{chazkel2007beyond}. The Baron called the lottery the \textit{jogo do bicho}, or the animal game, and it was well-received by the public. Eager to capitalise on that initial success, Drummond stated that visitors could buy tickets not only at the zoo, but also in many stores across Rio de Janeiro. What was once a small raffle soon became a large gambling market of its own.

A \textit{jogo do bicho} craze swept the whole city after independent sellers entered the marketplace. A network of street bookmakers, called \textit{bicheiros}, made the lottery available in every part of Rio by scalping tickets or promoting their own versions of the numbers game \citep[37]{chazkel2011laws}. The lottery became so widespread that Olavo Bilac, a major literary figure in nineteenth-century Brazil, summarised the situation as follows: `Today $[1985]$ in Rio de Janeiro, the game is everything. $[\dots]$  Nobody works! Everybody plays' \citep[43]{pacheco1957antologia}.\footnote{Unless otherwise noted, all translations from the Portuguese are my own.}

But this tolerant state of affairs did not last. Civil servants and police officers criminalised the \textit{jogo do bicho} on the grounds of `public safety', and in the late 1890s they launched a country-wide campaign against the lottery \citep{benatte2002jogos}. The campaign extended for several decades and received considerable support from the \textit{Companhia das Loterias Nacionaes do Brazil}, the National Lottery Company, a public-private partnership founded four years after, and perhaps motivated by, the creation of the animal game \citep[82]{da1999aguias}. The Brazilian government officially banned the \textit{jogo do bicho} in 1941 and the lottery remains illegal until this day.

Yet the game has survived. The \textit{jogo do bicho} has outlasted more than 30 Brazilian presidents and thrived under both military regimes and democratic governments \citep{jupiara2015poroes}. But more than an act of civil disobedience, the \textit{jogo do bicho} is a very successful capitalist enterprise \citep{labronici2014sorteio, magalhaes2005ganhou}. A recent study by Fundação Getúlio Vargas, a Brazilian think tank, affirmed that the \textit{jogo do bicho} earns from BRL 1.3 to BRL 2.8 billion per year (USD 400 to USD 850 million), what makes it the largest clandestine gambling game in the world \citep{huff2015jogos}. \citet[171]{schneider1996brazil} estimated that in the 1990s, the game furnished about 50,000 jobs in the Rio de Janeiro city alone, almost as many as the oil giant Petrobras in 2011 \citep{exame2016petrobras}.\footnote{In 1966, Time Magazine wrote that the \textit{jogo do bicho} was `the largest single industry in Latin America' and employed about 1\% of the total Brazilian workforce \citep{time1966bicho}.}

In this article I offer a rational choice interpretation of the \textit{jogo do bicho}.\footnote{I adopt a very broad definition of rationality in this paper. In contrast with stricter versions of rational choice theory, I assume here that individuals are not only constrained by formal and informal institutions, but they also have access only to imperfect information at the moment of their choices. Thus, my analysis employs a `thin' notion of rationality and a `thick' description of social institutions, where individual action can only be understood with reference to the social environment \citep[253]{boettke2001calculation}.} More specifically, I use an array of bibliographical sources to show how the game operators, commonly called \textit{bicheiros},\footnote{As I describe in the next section, the \textit{jogo do bicho} structure can be broadly divided into three occupations: the rich game financiers (\textit{banqueiros}), mid-level managers (\textit{gerentes}), and street bookmakers (\textit{bicheiros}). While there are important hierarchical differences among these groups, all members of the \textit{jogo} are collectively called \textit{bicheiros} in Brazil. I follow the same practice here, and only refer to their specific role within the animal game structure when necessary for clarification.} have developed unique strategies to solve collective action problems and maximise their political strategies. This does not mean that their tactics are morally defensible; \textit{bicheiros} regularly employ intimidation and corruption to achieve their goals. Nevertheless, I argue here that such strategies are effective, and while they seem counter-intuitive, they do address the long-term needs of the \textit{jogo do bicho} financiers.

Like any business manager, \textit{bicheiros} have to run their firm with low costs to increase profits. However, the fact that the \textit{jogo do bicho} is clandestine imposes additional difficulties to its managers. In particular, \textit{bicheiros} face two main challenges to keep the lottery running. First, they need to gather public support so that gamblers are not discouraged to engage in the lottery despite it being illegal. Second, \textit{bicheiros} also have to ensure that the state repression is not prohibitively costly to their business, otherwise they would be better off by shutting it down. I argue below that the \textit{bicheiros} have succeeded in both by using carefully-designed reputation strategies and employing costly signals to the communities they serve.

I use the case of Rio de Janeiro to illustrate how the \textit{jogo do bicho} has overcome the obstacles to its expansion. Rio is a particularly interesting case because in no other part of Brazil the game financiers established such an effective patronage network. \textit{Bicheiros} have sponsored political campaigns, financed cultural activities and football teams, and sometimes even run in local elections themselves. I discuss the ways by which the \textit{bicheiros} have exploited fragilities of the Brazilian political system to their advantage and how those practices have weakened Brazilian democracy.

My analysis discusses three strands of academic literature. First, this work contributes to the scholarship on extra-legal institutions, mainly to the literature on collective action within criminal organisations. For instance, \citet{gambetta1996sicilian} examines the strategies used by the Sicilian Mafia to settle disputes among their members and enforce rules in the areas they exercise control. \citet{leeson2009invisible,leeson2010pirational} affirms that pirate groups employed hard-to-fake signals to increase the profitability of their operations. \citet{skarbek2011governance,skarbek2012prison,skarbek2014social}, in turn, highlights the role of written and implicit norms in mitigating rent-seeking and coordinating productive activities in California prison gangs. I argue that \textit{bicheiros} have employed reputation strategies and provided club goods to enforce private contracts and foster trust in the community.

Second, this work relates to the literature on signalling theory and asymmetric information \citep[e.g.,][]{akerlof1970market,spence1973job}. I provide evidence that \textit{bicheiros} were aware of their social stigma, and as a response they devised signalling strategies to convey reliable information and reduce the uncertainty associated with clandestine markets. Their main tool to increase credibility was costly signalling \citep{gambetta2009codes,kimbrough2015commitment, schelling1960strategy}. \textit{Bicheiros} believed that by sacrificing their immediate interests they could gain a reputation of honesty that would benefit them in the long run.

Lastly, this work connects to the literature on state capture, which is one of the most important topics in public choice theory \citep{rose1978corruption,shleifer2002grabbing,tollison1982rent}. More specifically, I use the Brazilian case to illustrate how politicians and civil servants are co-opted by criminal groups, and how this collusion distorts the electoral process and benefits wealthy members of the \textit{jogo do bicho} network. \citet{queiroz1992carnaval} explored why \textit{bicheiros} turned into patrons of the Carnival's samba schools and affirmed that this influence gave them leverage over political authorities. \citet{misse2007illegal} investigated the links between bicheiros and police officers, and suggested that the illegal lottery had been the main cause of police corruption in Rio de Janeiro until the 1970s. In a similar vein, \citet{jupiara2015poroes} analyse the relationship between the \textit{jogo do bicho} and the military regime in Brazil (1964--1985). I supplement this literature by highlighting how asymmetrical information and rent-seeking behaviour offer convincing explanations to the issues presented above. Although those concepts have a long tradition in public choice, scholars have not applied those ideas to understand the dynamics of the \textit{jogo do bicho}. By doing so, I integrate seemingly contradictory historical facts into a single narrative that connects micro-level decisions to macro-level outcomes.

@@

The late nineteenth-century Brazil had four characteristics that explain the emergence of the \textit{jogo do bicho}. They are: 1) a growing urban population excluded from the formal labour market; 2) an inflow of immigrants whose extended family networks helped them engage in trade; 3) an expansion of the monetary supply in the first years of the republic (1880s--1890s); and 4) a judicial system that, albeit repressive, had only imperfect law enforcement. I discuss each of these elements below.

@@

As a result, large swathes of the urban population turned to the informal economy. As \citet[115]{chazkel2011laws} observes, there were few occupations available to lower-class women and foreigners in the 1890s, and a large number of poor workers became street vendors. The profession requires little technical skills and has low barriers of entry, but it can be profitable if vendors are able to quickly identify a growing demand for a particular product. The \textit{jogo do bicho} was one such case; not only the game had a surge in popularity in the 1890s, but it also offered a high rate of return. As the demand for lottery tickets grew, the \textit{jogo} comprised an important share of the Brazilian extra-legal economy.

Immigration also influenced the \textit{jogo do bicho} via social ties. Most foreigners who moved to Brazil came from countries, such as Portugal, Spain or Italy, where extended families were the basic form of social organisation \citep{lobo2001imigraccao, trento1989outro}. Family and neighbourhood networks created incentives for immigrants to establish trade relations and enforce cooperation through community responsibility systems \citep{roth2014prison}. Because of these particular social characteristics, in the 1890s foreigners were over-represented in the Brazilian trade in general \citep{mattos1991vadios, oliveira2001brasil} and in the \textit{jogo do bicho} in particular \citep{magalhaes2005ganhou, villar2008contravencao}. Although kinship bonds became less relevant over time, these links offered an important element of social cohesion in the \textit{jogo do bicho}'s formative years.

Next is the impact of expanded monetary supply. The abolition of slavery and the growing industrialisation of Brazil increased the amount of capital available in the country \citep{franco1987reformas, schulz2008financial}. Moreover, the 1888 Banking Act gave extra liquidity to local financial markets, what made credit more widely available in cities like São Paulo and Rio de Janeiro. Individuals received a temporary boost in personal income, a part of which they spent on leisure activities such as the \textit{jogo do bicho}. Moreover, the lottery attracted new entrants as it became more profitable, and in only a few years similar versions of the animal game were available throughout Brazil \citep[79]{da1999aguias}.

>>>>

\begin{document}
\maketitle
\begin{abstract}
\noindent
This article presents a rational choice account of Brazil's \textit{jogo do bicho} (``animal game''), possibly the largest illegal lottery game in the world. In over 120 years, the \textit{jogo do bicho} has grown from a local raffle to a multimillion-dollar business, and the game has played a major role in Brazil's cultural and political life. My analysis of the \textit{jogo do bicho} has three goals. First, I examine the mechanisms that fostered the lottery's notable growth outside the boundaries of Brazilian law. Second, I investigate how the animal game financiers combine costly signals and selective incentives to induce cooperation from members of the community. Lastly, I discuss the relationship between the lottery sponsors and Brazilian representatives, particularly how the \textit{bicheiros} exploit the fragmentation of Brazil's political system to advance their long-term interests.

@@

In 1892, Baron João Batista de Viana Drummond came up with a new idea to fund his cash-strapped zoo. Situated in a quiet neighbourhood in the north of Rio de Janeiro, the \textit{Jardim Zoológico}, or Zoological Garden, hosted a variety of exotic species and offered breath-taking views of the city, but it lacked visitors. An experienced businessman, Drummond realised the zoo would have to provide other kinds of entertainment to keep itself afloat. One of his plans seemed particularly promising: a lottery raffle.

The rules were straightforward. In the morning, the baron would choose one animal from a list of 25 beasts and put its picture inside a wooden box at the zoo's entrance. Visitors who wanted to join the raffle received a ticket bearing the stamp of one of those 25 animals.\footnote{At first, the zoo staff distributed the tickets at random, so the game consisted in a simple raffle. However, the zoo soon allowed participants to choose the animals they preferred. This small change made the game considerably more appealing as it introduced an element of divination to the \textit{jogo} \citep[71--74]{da1999aguias}.} The lucky winner would take home a prize worth 20 times the ticket price, an amount higher than a carpenter's monthly wage \citep[542]{chazkel2007beyond}. The baron called the lottery the \textit{jogo do bicho}, or the animal game, and it was well-received by the public. Eager to capitalise on that initial success, Drummond stated that visitors could buy tickets not only at the zoo, but also in many stores across Rio de Janeiro. What was once a small raffle soon became a large gambling market of its own.

A \textit{jogo do bicho} craze swept the whole city after independent sellers entered the marketplace. A network of street bookmakers, called \textit{bicheiros}, made the lottery available in every part of Rio by scalping tickets or promoting their own versions of the numbers game \citep[37]{chazkel2011laws}. The lottery became so widespread that Olavo Bilac, a major literary figure in nineteenth-century Brazil, summarised the situation as follows: ``Today $[1985]$ in Rio de Janeiro, the game is everything. $[\dots]$  Nobody works! Everybody plays''  \citep[43]{pacheco1957antologia}.\footnote{Unless otherwise noted, all translations from the Portuguese are my own.}

But this tolerant state of affairs did not last. Civil servants and police officers criminalised the \textit{jogo do bicho} on the grounds of public safety, and in the late 1890s they launched a country-wide campaign against the lottery \citep{benatte2002jogos}. The campaign extended for several decades and received considerable support from the \textit{Companhia das Loterias Nacionaes do Brazil}, the National Lottery Company, a public-private partnership founded four years after, and perhaps motivated by, the creation of the animal game \citep[82]{da1999aguias}. The Brazilian government officially banned the \textit{jogo do bicho} in 1941 and the lottery remains illegal to this day.

Yet the game has survived. The \textit{jogo do bicho} has outlasted more than 30 Brazilian presidents and thrived under both military regimes and democratic governments \citep{jupiara2015poroes}. But more than an act of civil disobedience, the \textit{jogo do bicho} is a very successful capitalist enterprise \citep{labronici2014sorteio, magalhaes2005ganhou}. A recent study by Fundação Getúlio Vargas, a Brazilian think tank, affirmed that the \textit{jogo do bicho} earns from BRL 1.3 to BRL 2.8 billion per year (USD 400 to USD 850 million), making it the largest clandestine gambling game in the world \citep{huff2015jogos}. \citet[171]{schneider1996brazil} estimated that in the 1990s, the game furnished about 50,000 jobs in Rio de Janeiro, almost as many as the oil giant Petrobras in 2011 \citep{exame2016petrobras}.\footnote{In 1966, Time Magazine wrote that the \textit{jogo do bicho} was ``the largest single industry in Latin America'' and employed about 1\% of the total Brazilian workforce \citep{time1966bicho}.}

In this article I offer a rational choice interpretation of the \textit{jogo do bicho}.\footnote{I adopt a very broad definition of rationality in this paper. In contrast with stricter versions of rational choice theory, I assume here that individuals are not only constrained by formal and informal institutions, but they also have access only to imperfect information at the moment of their choices. Thus, my analysis employs a ``thin'' notion of rationality and a `thick' description of social institutions, where individual action can only be understood with reference to the social environment \citep[253]{boettke2001calculation}.} More specifically, I use an array of bibliographical sources to show how the game operators, commonly called \textit{bicheiros},\footnote{As I describe in the next section, the \textit{jogo do bicho} structure can be broadly divided into three occupations: the rich game financiers (\textit{banqueiros}), mid-level managers (\textit{gerentes}), and street bookmakers (\textit{bicheiros}). While there are important hierarchical differences among these groups, all members of the \textit{jogo} are collectively called \textit{bicheiros} in Brazil. I follow the same practice here, and only refer to their specific role within the animal game structure when necessary for clarification.} have developed unique strategies to solve collective action problems and to maximise their political strategies. This does not mean their tactics are morally defensible; \textit{bicheiros} regularly employ intimidation and corruption to achieve their goals. Nevertheless, I argue here that such strategies are effective, and while they seem counter-intuitive, they do address the long-term needs of the \textit{jogo do bicho} financiers.

Like any business manager, \textit{bicheiros} have to run their firm with low costs to increase profits. However, the fact that the \textit{jogo do bicho} is clandestine imposes additional difficulties to its managers. In particular, \textit{bicheiros} face two main challenges to keep the lottery running. First, they need to gather public support so gamblers are not discouraged to engage in the lottery despite it being illegal. Second, \textit{bicheiros} also have to ensure the state repression is not prohibitively costly to their business; otherwise they would be better off shutting it down. I argue below that the \textit{bicheiros} have succeeded in both by using carefully-designed reputation strategies and employing costly signals to the communities they serve.

I use the case of Rio de Janeiro to illustrate how the \textit{jogo do bicho} has overcome the obstacles to its expansion. Rio is a particularly interesting case because in no other part of Brazil is the game the financiers established such an effective patronage network. \textit{Bicheiros} have sponsored political campaigns, financed cultural activities and football teams, and sometimes even run in local elections themselves. I discuss the ways by which the \textit{bicheiros} have exploited fragilities of the Brazilian political system to their advantage and how those practices have weakened Brazilian democracy.

My analysis discusses three strands of academic literature. First, this work contributes to the scholarship on extra-legal institutions, mainly to the literature on collective action within criminal organisations. For instance, \citet{gambetta1996sicilian} examines the strategies used by the Sicilian Mafia to settle disputes among their members and enforce rules in the areas they exercise control. \citet{leeson2009invisible,leeson2010pirational} affirms that pirate groups employed hard-to-fake signals to increase the profitability of their operations. \citet{skarbek2011governance,skarbek2012prison,skarbek2014social}, in turn, highlights the role of written and implicit norms in mitigating rent-seeking behaviour and in coordinating productive activities in California prison gangs. I argue that \textit{bicheiros} have employed reputation strategies and provided club goods to enforce private contracts and to foster trust in the community.

Second, this work relates to the literature on signalling theory and asymmetric information \citep[e.g.,][]{akerlof1970market,spence1973job}. I provide evidence that \textit{bicheiros} were aware of their social stigma, and as a response they devised signalling strategies to convey reliable information and to reduce the uncertainty associated with clandestine markets. Their main tool to increase credibility was costly signalling \citep{gambetta2009codes,kimbrough2015commitment, schelling1960strategy}. \textit{Bicheiros} believed that by sacrificing their immediate interests they could gain a reputation of honesty that would benefit them in the long run.

Lastly, this work connects to the literature on state capture, which is one of the most important topics in public choice theory \citep{rose1978corruption,shleifer2002grabbing,tollison1982rent}. More specifically, I use the Brazilian case to illustrate how politicians and civil servants are co-opted by criminal groups, and how this collusion distorts the electoral process and benefits wealthy members of the \textit{jogo do bicho} network. \citet{queiroz1992carnaval} explored why \textit{bicheiros} turned into patrons of the Carnival's samba schools and affirmed this influence gave them leverage over political authorities. \citet{misse2007illegal} investigated the links between bicheiros and police officers, and suggested the illegal lottery had been the main cause of police corruption in Rio de Janeiro until the 1970s. In a similar vein, \citet{jupiara2015poroes} analyse the relationship between the \textit{jogo do bicho} and the military regime in Brazil (1964--1985). I supplement this literature by highlighting how asymmetrical information and rent-seeking behaviour offer convincing explanations to the issues presented above. Although those concepts have a long tradition in public choice, scholars have not applied those ideas to understand the dynamics of the \textit{jogo do bicho}. By doing so, I integrate seemingly contradictory historical facts into a single narrative that connects micro-level decisions to macro-level outcomes.

@@

The late nineteenth-century Brazil had four characteristics that explain the emergence of the \textit{jogo do bicho}: 1) a growing urban population excluded from the formal labour market; 2) an inflow of immigrants whose extended family networks helped them engage in trade; 3) an expansion of the monetary supply in the first years of the republic (1880s--1890s); and 4) a judicial system that, albeit repressive, had only imperfect law enforcement. I discuss each of these elements below.

@@

As a result, large swathes of the urban population turned to the informal economy. As \citet[115]{chazkel2011laws} observes, there were few occupations available to lower-class women and foreigners in the 1890s, and a large number of poor workers became street vendors. The profession requires little technical skills and has low barriers of entry, but it can be profitable if vendors are able to quickly identify a growing demand for a particular product. The \textit{jogo do bicho} was one such case; the game had a surge in popularity in the 1890s, and it also offered a high rate of return. As the demand for lottery tickets grew, the \textit{jogo} comprised an important share of the Brazilian extra-legal economy.

Immigration also influenced the \textit{jogo do bicho} via social ties. Most foreigners who moved to Brazil came from countries such as Portugal, Spain or Italy, where extended families were the basic form of social organisation \citep{lobo2001imigraccao, trento1989outro}. Family and neighbourhood networks created incentives for immigrants to establish trade relations and to enforce cooperation through community responsibility systems \citep{roth2014prison}. Because of these particular social characteristics, in the 1890s foreigners were over-represented in the Brazilian trade in general \citep{mattos1991vadios, oliveira2001brasil} and in the \textit{jogo do bicho} in particular \citep{magalhaes2005ganhou, villar2008contravencao}. Although kinship bonds became less relevant over time, these links offered an important element of social cohesion in the \textit{jogo do bicho}'s formative years.

Next is the impact of expanded monetary supply. The abolition of slavery and the growing industrialisation of Brazil increased the amount of capital available in the country \citep{franco1987reformas, schulz2008financial}. Moreover, the 1888 Banking Act gave extra liquidity to local financial markets, which made credit more widely available in cities like São Paulo and Rio de Janeiro. Individuals received a temporary boost in personal income, a part of which they spent on leisure activities such as the \textit{jogo do bicho}. Moreover, the lottery attracted new entrants as it became more profitable, and in only a few years similar versions of the animal game were available throughout Brazil \citep[79]{da1999aguias}.

The last necessary condition for the emergence of the \textit{jogo do bicho} is weak law enforcement. \citet[69--100]{chazkel2011laws} notes that until the 1940s, police district chiefs operated within a large margin of discretion, so official repression against bookmakers was notably inconsistent. In the early years of \textit{jogo do bicho}, lottery ``bankers'' were allowed to operate virtually free from police interference, which surely collaborated to the game's rapid initial expansion \citep[544]{chazkel2007beyond}. Prosecution against the \textit{bicheiros} hardened in 1917 after the promulgation of the Civil Code, and in 1941 the animal game was banned. Five years later, the federal government declared that all games of chance were illegal in Brazil.\footnote{The 1946 decree stated that gambling was `harmful to morality and the good customs', hence `[\dots] the repression against games of chance [was] an imperative of the universal consciousness'.} Recent estimations show that the prohibition of the \textit{jogo do bicho} prevented the state from earning up to BRL 20 billion (USD 6 billion) per year in expected taxation revenues, aside from the subjective utility losses for players \citep{fsp2016legalizarbicho}.
